### 2012-11-22 paper/isca40format/Introduction.tex
before reorging scheduler etc

Large cache-coherent shared-memory multicores are likely to use protocols which
are directory-based, hierarchical, and implemented using packet-switched networks.  Even
though such protocols have been around for two decades (see, for example,
DASH\cite{DASH}, Alewife\cite{Anant}) correct design and implementation of
such protocols remains a challenge. It is difficult to convince oneself that
all corner cases have been handled adequately without a formal proof. And yet
formal proofs of real implementations are practically non-existent. One can't
settle for conservative choices in cache-coherence protocol designs because
performance (\eg the number of messages needed to maintain coherence, the
buffer requirements, etc) is as important as correctness. Consequently proving
a simple protocol for a small number of caches, under unbounded buffer
assumption is of limited use in gaining confidence in the correctness of a real
protocol.

@@

Every cache, unless it is an L1 cache, has a directory which contains
information about the state of its children caches. The cache state-changes to
maintain coherence are orchestrated by passing requests and responses between
the caches using the local directory information.  Unlike bus-based protocols
which lock the bus till all the required state transitions are done thereby
ensuring atomicity of state transitions, in a distributed protocol,
requests and responses are split leading to transient states. This creates
scenarios not present in bus-based protocols, which can lead to coherence
violations or deadlocks, if the protocol is not designed or implemented
carefully.

@@

In this paper, we present a substantially easier method to design cache
coherence protocols for such systems. The user specifies a protocol for a cache
hierarchy as sequential methods to process requests. It is not distributed, so the protocol procedure can simultaneously or \emph{atomically}
examine and change the state of all the caches in the system. We then
systematically transform such a protocol description into a distributed protocol. The resulting protocol is presented as a collection
of cooperating sequential methods which operate on ``cache objects''.  These
sequential methods communicate with each other via blocking \emph{send} and
\emph{receive} commands and thus can get suspended. Our transformation is based
on preserving some invariants which guarantee that each request eventually gets
a response and thus, our threads can never wait indefinitely.  It has been
shown using the Coq theorem prover \cite{}, that the distributed protocol we
derive is correct.  Both the transformation procedure and the proof are
parameterized by the number of cache levels and the number of caches in each
level of the hierarchy as well as by a predicate which involves a local
``compatibility of state'' test. A discussion of the formal proof is beyond the
scope of this paper.

It should be noted that the generality of our procedure comes from the fact
that we do not assign any meaning to a coherence state beyond upgrades and
downgrades. Thus, how or why a particular cache state implies that a particular
address may not be present in any other cache stems from protocol specific
``compatibility of state'' check.

The main contributions of this paper are: 1. An automatic procedure to refine a
centralized MSI protocol into protocols for
distributed, hierarchical, cache system with packet-switching; 2. Specification
of the requirements that ensure the correctness of the generated protocol;
3. A discussion of the invariants that simplifies the proof of correctness of
the protocol considerably;
4. Giving a provable lower bound on the number of network virtual channels and
other resources to avoid deadlocks; and 5. Showing that our procedure does not
introduce any overhead over direct implementation of these protocols. We think
that our method of designing protocols substantially reduces the required
effort in implementing provably correct protocols.

\paragraph{Paper organization:} Section \ref{sec:GlobalMsi} describes the MSI
protocol for a hierarchy of caches using nonsuspensive sequential methods. We
use it a specification for proving the correctness of distributed MSI
protocols. Section \ref{sec:DistributedMsi} systematically transforms the
protocol in Section \ref{sec:GlobalMsi} into a distributed protocol. Section
\ref{sec:network} specifies the network, resource and other requirements that
are needed for the distributed MSI protocol. We also give the minimum virtual
channel requirements to ensure the requirements specified in Section
\ref{sec:network}. In Section \ref{sec:implementation}, we show that these formally verified family of protocols
require no extra state bits, nor extra message transfers compared to other
published implementations. Section \ref{sec:properties} gives the local
properties of the distributed protocol that we rely on to easily establish its
correctness, and show how it can be generalized to design other protocols like MOSI.
%It also lists core properties of the distributed protocol that are needed to establish the invariants.
In Section \ref{sec:related}, we present some of the related work and finally
in Section \ref{sec:conclusion}, we present some directions for future work.

>>>>

Large cache-coherent shared-memory multicores are likely to use protocols which
are directory-based, hierarchical, and implemented using packet-switched networks.  Even
though such protocols have been around for two decades (see, for example,
\cite{DASH, Anant, c1, c2, c3, c4, c5}) correct design and implementation of
such protocols remains a challenge. It is difficult to convince oneself that
all corner cases have been handled adequately without a formal proof. And yet
formal proofs of real implementations are practically non-existent. One can't
settle for conservative choices in cache-coherence protocol designs because
performance (\eg the number of messages needed to maintain coherence, the
buffer requirements, etc) is as important as correctness. Consequently proving
a simple protocol for a small number of caches, under unbounded buffer
assumption is of limited use in gaining confidence in the correctness of a real
protocol.

@@

Every cache, unless it is an L1 cache, has a directory which contains
information about the state of its children caches. The changes in the cache states to
maintain coherence are orchestrated by passing requests and responses between
the caches using the local directory information.  Unlike bus-based protocols
which conceptually lock the bus till all the required state transitions are done thereby
ensuring atomicity of state transitions, in a distributed protocol,
requests and responses are split leading to transient states. This creates
scenarios not present in bus-based protocols, which can lead to coherence
violations or deadlocks, if the protocol is not designed or implemented
carefully.

@@

In this paper, we present a substantially easier method to design cache
coherence protocols for such systems. The user specifies a protocol for a cache
hierarchy as sequential methods to process requests. It is not distributed, so the protocol procedure can simultaneously or \emph{atomically}
examine and change the state of all the caches in the system. We then
systematically transform such a protocol description into a distributed protocol. The resulting protocol is presented as a collection
of cooperating sequential methods which operate on ``cache objects''.  These
sequential methods communicate with each other via blocking \emph{send} and
\emph{receive} commands and thus can get suspended. Our transformation is based
on preserving some invariants which guarantee that each request eventually gets
a response and thus, our threads can never wait indefinitely.  It has been
shown using the Coq theorem prover \cite{coq}, that the distributed protocol we
derive is correct.  Both the transformation procedure and the proof are
parameterized by the number of cache levels and the number of caches in each
level of the hierarchy as well as by a predicate which involves a local
``compatibility of state'' test. A discussion of the formal proof is beyond the
scope of this paper.

It should be noted that the generality of our procedure comes from the fact
that we do not assign any meaning to a coherence state beyond upgrades and
downgrades. Thus, how or why a particular cache state implies that a particular
address may not be present in any other cache stems from a protocol specific
``compatibility of state'' check.

The main contributions of this paper are: 1. An automatic procedure to refine a
centralized MSI protocol into a protocol for
distributed, hierarchical cache system with packet-switching; 2. Specification
of requirements for network, resources and local scheduling to ensure the correctness of the generated protocol;
3. A provable lower bound on the number of virtual channels and storage
 to avoid deadlocks and showing that these requirements do not introduce any overhead over published implementations. We think
that our method of designing protocols substantially reduces the required
effort in implementing provably correct distributed protocols.

\paragraph{Paper organization:} Section \ref{sec:GlobalMsi} describes the MSI
protocol for a hierarchy of caches using nonsuspensive sequential methods. We
use it a specification for proving the correctness of distributed MSI
protocols. Section \ref{sec:DistributedMsi} systematically transforms the
protocol in Section \ref{sec:GlobalMsi} into a distributed protocol. We specify
the assumptions of the network, and other requirements that are needed for the
distributed MSI protocol. We also give the minimum virtual channel requirements
to ensure that the network assumptions are satisfied. In Section
\ref{sec:implementation}, we show that these formally verified family of
protocols require no extra state bits, nor extra message transfers compared to
other published implementations. Section \ref{sec:properties} gives the local
properties of the distributed protocol that we rely on to easily establish its
correctness, and show how it can be generalized to design other protocols like
MOSI.
%It also lists core properties of the distributed protocol that are needed to
%establish the invariants.
In Section \ref{sec:related}, we present some of the related work and finally
in Section \ref{sec:conclusion}, we present some conclusions.

### 2012-11-22 paper/isca40format/Invariants.tex
before reorging scheduler etc

\section{Local properties that ensure global invariants}
\label{sec:properties}

In this section, we will show some of the core properties that are extracted
from the requirements spelt out in Section \ref{sec:network} as well as the
implementation of the distributed protocol's auxiliary functions given in Figure
\ref{realistic}. These properties can be independently used to prove the
invariants presented in Section \ref{sec:DistributedMsi} (Invariants
\ref{conservative} to \ref{cpSame}). We do not present the proof here, but give
some intuition. Figure \ref{sendReq} lists these local properties.

@@

\begin{figure}\small
\begin{inv}
A response \Resp{p}{c}{a}{x}, where $p = c.parent$, can be sent from $p$ only
if
\begin{enumerate}
\item a request \Req{c}{p}{a}{z} was received from $c$,
\item $p$ is not waiting for any response for address $a$ from $c$, and
\item $x > y$, where $p.dir[c][a] = y$ just before sending the response
\end{enumerate}
\label{pSendRespPre}
\end{inv}
\begin{inv}
A response \Resp{c}{p}{a}{x}, where $p = c.parent$, can be sent from $c$ only
if $x < y$, where $c.state[a] = y$ just before sending the
response\label{cSendRespPre1}
\end{inv}
\begin{inv}
If $c$ has sent a request \Req{c}{p}{a}{x}, then $c$ can send a response
\Resp{c}{p}{a}{y} only on receiving a request from $p$ for address $a$.
\label{cSendRespPre2}
\end{inv}
%\caption{Local-Properties for sending responses}
%\label{sendResp}
%\end{figure}
%
%\begin{figure}\small
\begin{inv}
If a response \Resp{p}{c}{a}{x} is sent from $p$, then $p.dir[c][a] \gets x$\label{cSendRespPost}
\end{inv}
\begin{inv}
If \Resp{c}{p}{a}{x} is sent $c.state[a] \gets x$\label{pSendRespPost}
\end{inv}
\begin{inv}
On receiving a response \Resp{c}{p}{a}{x}, where $p = c.parent$, $p.dir[c][a]
\gets x$\label{pRecvResp}
\end{inv}
\begin{inv}
On receiving a response \Resp{p}{c}{a}{x}, where $p = c.parent$, $c.state[c]
\gets x$\label{cRecvResp}
\end{inv}
\begin{inv}
$c.state[a]$ can change only on $c$ sending a response to or receiving a
response from $c.parent$\label{cState}
\end{inv}
\begin{inv}
$p.dir[c][a]$ can change only on $p$ sending or receiving a response from
its child $c$\label{pState}
\end{inv}
%\caption{Local-Properties for state changes}
%\label{stateChange}
%\end{figure}
%
%\begin{figure}\small
\begin{inv}
\Req{p}{c}{a}{x}, where $p = c.parent$, can be sent only if $x < p.dir[c][a]$\label{pSendReqPre}
\end{inv}
\begin{inv}
If \Req{p}{c}{a}{x}, where $p = c.parent$, is received, if $x \ge c.state[a]$,
then the request is dropped\label{pSendReqPost}
\end{inv}
\begin{inv}
\Req{c}{p}{a}{x}, where $p = c.parent$, can be sent only if $x > c.state[a]$\label{cSendReqPre}
\end{inv}
\begin{inv}
If \Req{c}{p}{a}{x}, where $p = c.parent$, is received, it $x \le p.state[a]$,
then the request is dropped\label{cSendReqPost}
\end{inv}
\begin{inv}
A request can be sent by a cache node $n$ for address $n$ to another cache node $m$ only if
no $n$ is not waiting for a response for address $a$ from $m$.
\label{nodoublereq}
\end{inv}
\caption{Local-Properties useful in proving the correctness of MSI protocol}
\label{sendReq}
\end{figure}

@@

In a hierarchical setting for the MOSI protocol, any cache can send data
with the response granting permissions to the requesting L1 cache, instead of
going through the hierarchy.  On a load request, L1 cache $c$ sends an
upgrade-to-$O$ request, instead of an upgrade-to-$S$ request. The request keeps
going up the hierarchy, sending an upgrade-to-$O$ request to each of the parent
on the way, till it reaches a cache $p$ which has that line in $M$ or $O$
state. If some other child is in $O$ or $M$ state, this node sends a
upgrade-$c$-to-$O$ in addition to a downgrade-to-$S$ request to that cache.
This upgrade-$c$-to-$O$ + downgrade-to-$S$ keeps propogating down the tree till
it reaches a node $n$ whose children are all in a state $< O$. $n$ then
downgrades its state, forwards the upgrade-to-$O$ response to $c$ and
simultaneously sends a fwd-and-downgrade-to-$S$ response to its parent, which
keeps forwarding that till it reaches $p$. If $c$ gets a store request, it will
send an upgrade-to-$M$ request, and $p$ would send a downgrade-to-$I$ request
instead, to its children.

@@

This protocol obeys all the local properties and the requirements mentioned in
this section except that a request sent to a cache might overtake a response
forwarded to that cache. To avoid this, once the forwarding message is
delivered to the intended destination ($c$ in our case), then $c$ sends back an
ACK. $c$'s parent is allowed to to start sending downgrade requests for that
address only after receiving this ACK. This ACK is sent back to each of the
parents till it reaches $p$, which allows each of these caches to send
downgrade requests for that address towards $c$.  This satisfies all the
requirements of Section \ref{sec:network} and the local properties given in
Figure \ref{sendReq}, thereby automatically being correct. The reasoning for
its correctness gets localized to checking if the ordering Requirement
\ref{reqNoOvertakeResp} is maintained.

>>>>

\section{Intuition underlying the automatic implementation}
\label{sec:properties}

In this section, give the intuition on how we ensured that the implementation of
the hierarchical MSI protocol is correct.

The most important invariant that the generated protocol maintains between the
caches and its parent is as follows:
\begin{theorem}
Conservative: For two nodes $p, c$, where $p = c.parent$, $\forall a,
p.dir[c][a] \ge c.state[a]$.  \label{conservative}
\end{theorem}

This ensures that if a cache makes a decision about compatibility of states of
its children using the local directory information, then it does not violate the
compatibility of the children's real states. This essentially ensures that
reading the directory is sufficient to know the value of a cache's state.

Regarding the \send{} and \receive{} operations, the following
invariants are very critical to ensure that a thread completes. If they are not
true, then a thread might block waiting for a response, or being unable to send a
message, leading to a deadlock. Our generated protocol maintains this.

\begin{theorem}
It a thread has to send a message, it will eventually be able to send the
message.
\label{canSend}
\end{theorem}

\begin{theorem}
If a cache sends an upgrade-to-$x$ request to its parent, it will eventually
get an upgrade-to-$y$ response from the parent where $y \ge x$.
\label{willRecv}
\end{theorem}

\begin{theorem}
If a cache sends an downgrade-to-$x$ request to one of its children, it will
eventually get a downgrade-to-$y$ response from the child where $y \le x$.
\label{willRecv2}
\end{theorem}

Another scenario where a thread can potentially get suspended is if handling of
a request requires eviction of another cache line. The thread would get
suspended if all the lines that can be allocated for this request are being 
accessed by other suspended threads. But the generated protocol still guarantees
the following invariant.

\begin{theorem}
If a thread is waiting for a line to be allocated, \ie for any address to become
available for eviction, then some address will eventually become available for
eviction
\label{evictDead}
\end{theorem}

It this is not true, a thread that is waiting for being allocated a free line
may never proceed, leading to a deadlock.

If a cache receives a response message, it must know whether or not a) it will
be receiving data, and b) whether it has to update the address with the received
data, in case it does receive data. The first part can be indicated by the
message itself, but the second part has to be decided only after knowing the
state of the cache which sent the response. In this regard, the following two
invariants are very useful.

\begin{theorem}
If $p$ sends \Resp{p}{c}{a}{x} to $c$, where $p = c.parent$, then $p.dir[c][a]$
just before $p$ sends the response is the same as $c.state[a]$ just before $c$
receives the response.
\label{pcSame}
\end{theorem}

\begin{theorem}
If $c$ sends \Resp{c}{p}{a}{x} to $p$, where $p = c.parent$, then $c.state[a]$
just before $c$ sends the response is the same as $p.dir[c][a]$ just before $p$
receives the response.
\label{cpSame}
\end{theorem}

We will illustrate the need for Invariants \ref{pcSame}, \ref{cpSame} and
\ref{conservative} using an example of a 2-level system containing two L1
caches $c_1$ and $c_2$ with a shared L2 cache $p$.

Invariant \ref{pcSame} and \ref{cpSame} are important to make sure that the
data transfers happen correctly. Let $p.dir[c_1][a] = S$, $p.dir[c_2][a] = I$,
$c_1.state[a] = S$, and $c_2.state[a] = I$. $c_2$ gets a store request, and
sends an upgrade-to-$M$ request to $p$. $p$ receives the request, but assumes that
$c_1$ already has data for $a$ since $p.dir[c_1][a] = S$. Invariant
\ref{pcSame} is violated. So, it does not transfer the data to $c_1$, resulting
in $c_1$ waiting for data, and hence deadlock. A dual condition can be
constructed similarly, in which $c_1$ downgrades its state for address $a$ from
$S$ to $I$, and the parent assumes that $c_1$ is in state $M$, and keeps
waiting for data.

Invariant \ref{conservative} is important because a parent node consults its
directory to decide the children it needs to send a downgrade request to.  Let
$p.dir[c_1][a] = S$, $c_1.state[a] = S$, $p.dir[c_2][a] = I$ but
$c_2.state[a] = S$. If $c_1$ gets a store request from the processor and sends
an upgrade-to-$M$ request to $p$, $p$ will assume that $c_2$'s state for address $a$
is also $I$ and hence send an upgrade-to-$M$ response to $c_1$. But since the real
state of  $c_2$ for address $a$ is $S$, it will break Single-writer Invariant
\ref{singleWriter}.

Thus, if all the invariants above (\ref{conservative}, to \ref{cpSame}) hold,
then the threads given in Figure \ref{realistic} cannot deadlock because it
will always be able to a) send a request, b) send back a response, c) get back
a response for a request it sent, d) receive data when required. Though the
threads executing in a cache are actually suspensive, these invariants give an

We will show some of the core properties that are extracted
from the requirements spelt out in Section \ref{sec:DistributedMsi} and the
specification of the protocol (Figures \ref{msi-template} and
\ref{realistic}). These properties alone can be used to prove the above
invariants (Invariants \ref{conservative} to \ref{cpSame}). We do not present
the proof here, but give some intuition. Figure \ref{sendReq} lists these local
properties.

@@

\begin{figure}\small
\begin{inv}
A response \Resp{p}{c}{a}{x}, where $p = c.parent$, can be sent from $p$ only
if
\begin{enumerate}
\item a request \Req{c}{p}{a}{z} was received from $c$,
\item $p$ is not waiting for any response for address $a$ from $c$, and
\item $x > y$, where $p.dir[c][a] = y$ just before sending the response
\end{enumerate}
\label{pSendRespPre}
\end{inv}
\begin{inv}
A response \Resp{c}{p}{a}{x}, where $p = c.parent$, can be sent from $c$ only
if $x < y$, where $c.state[a] = y$ just before sending the
response\label{cSendRespPre1}
\end{inv}
\begin{inv}
If $c$ has sent a request \Req{c}{p}{a}{x}, then $c$ can send a response
\Resp{c}{p}{a}{y} only on receiving a request from $p$ for address $a$.
\label{cSendRespPre2}
\end{inv}
\caption{Local-Properties for sending responses}
\label{sendResp}
\end{figure}

The properties in Figure \ref{sendResp} restrict when responses can be sent by
 a cache (either to its children or to its parent).

\begin{figure}\small
\begin{inv}
If a response \Resp{p}{c}{a}{x} is sent from $p$, then $p.dir[c][a] \gets x$\label{cSendRespPost}
\end{inv}
\begin{inv}
If \Resp{c}{p}{a}{x} is sent $c.state[a] \gets x$\label{pSendRespPost}
\end{inv}
\begin{inv}
On receiving a response \Resp{c}{p}{a}{x}, where $p = c.parent$, $p.dir[c][a]
\gets x$\label{pRecvResp}
\end{inv}
\begin{inv}
On receiving a response \Resp{p}{c}{a}{x}, where $p = c.parent$, $c.state[c]
\gets x$\label{cRecvResp}
\end{inv}
\begin{inv}
$c.state[a]$ can change only on $c$ sending a response to or receiving a
response from $c.parent$\label{cState}
\end{inv}
\begin{inv}
$p.dir[c][a]$ can change only on $p$ sending or receiving a response from
its child $c$\label{pState}
\end{inv}
\caption{Local-Properties for state changes}
\label{stateChange}
\end{figure}

The properties in Figure \ref{stateChange} specify exactly when the coherence state of
a line (both directory and state) changes.

\begin{figure}\small
\begin{inv}
\Req{p}{c}{a}{x}, where $p = c.parent$, can be sent only if $x < p.dir[c][a]$\label{pSendReqPre}
\end{inv}
\begin{inv}
If \Req{p}{c}{a}{x}, where $p = c.parent$, is received, if $x \ge c.state[a]$,
then the request is dropped\label{pSendReqPost}
\end{inv}
\begin{inv}
\Req{c}{p}{a}{x}, where $p = c.parent$, can be sent only if $x > c.state[a]$\label{cSendReqPre}
\end{inv}
\begin{inv}
If \Req{c}{p}{a}{x}, where $p = c.parent$, is received, it $x \le p.state[a]$,
then the request is dropped\label{cSendReqPost}
\end{inv}
\begin{inv}
A request can be sent by a cache node $n$ for address $n$ to another cache node $m$ only if
node $n$ is not waiting for a response for address $a$ from $m$.
\label{nodoublereq}
\end{inv}
\caption{Local-Properties useful in proving the correctness of MSI protocol}
\label{sendReq}
\end{figure}

@@

In a hierarchical setting for the MOSI protocol, any cache can send data
with the response granting permissions to the requesting L1 cache, instead of
going through the hierarchy.  On a load request, L1 cache $c$ sends an
upgrade-to-$O$ request, instead of an upgrade-to-$S$ request. The request keeps
going up the hierarchy, sending an upgrade-to-$O$ request to each of the parent
on the way, till it reaches a cache $p$ which has that line in $M$ or $O$
state. If some other child is in $O$ or $M$ state, this node sends a
upgrade-$c$-to-$O$ in addition to a downgrade-to-$S$ request to that cache.
This upgrade-$c$-to-$O$ + downgrade-to-$S$ keeps propagating down the tree till
it reaches a node $n$ whose children are all in a state $< O$. $n$ then
downgrades its state, forwards the upgrade-to-$O$ response to $c$ and
simultaneously sends a fwd-and-downgrade-to-$S$ response to its parent, which
keeps forwarding that till it reaches $p$. If $c$ gets a store request, it will
send an upgrade-to-$M$ request, and $p$ would send a downgrade-to-$I$ request
instead, to its children.

@@

This protocol obeys all the local properties and the requirements mentioned in
this section except that a request sent to a cache might overtake a response
forwarded to that cache. To avoid this, once the forwarding message is
delivered to the intended destination ($c$ in our case), then $c$ sends back an
ACK. $c$'s parent is allowed to start sending downgrade requests for that
address only after receiving this ACK. This ACK is sent back to each of the
parents till it reaches $p$, which allows each of these caches to send
downgrade requests for that address towards $c$.  This satisfies all the
requirements of Section \ref{sec:DistributedMsi} and the local properties given in
Figures \ref{sendResp}, \ref{stateChange} and \ref{sendReq}, thereby automatically being correct. The reasoning for
its correctness gets localized to checking if the ordering Requirement
\ref{reqNoOvertakeResp} is maintained.

The protocols can also be extended in a different dimension. The proof is
oblivious to the states, so any downgrade of a state is treated the same. This
means that a cache line can downgrade from $M$ to $S$ state instead of going all
the way down to $I$ state. This flexibility can potentially be exploited, and we
will study its effects in the future.

### 2012-11-22 paper/isca40format/Msi.tex
before arvind is going to re-org
before reorging scheduler etc

\begin{wrapfigure}{r}{.6\linewidth}
\begin{subfigure}{.55\linewidth}
\centering
\begin{tabular}{|ccc|}
\hline
$M$ & $\rightarrow$ & $I$\\
$S$ & $\rightarrow$ & $S$\\
$I$ & $\rightarrow$ & $I$\\
\hline
\end{tabular}
\caption{$maxCompatible$}
\label{toCompat}
\end{subfigure}
\begin{subfigure}{.24\linewidth}
\centering
\begin{tabular}{|c|c|}
\hline
$M$ & $True$\\
$S$ & $False$\\
$I$ & $False$\\
\hline
\end{tabular}
\caption{$isModified$}
\label{isModified}
\end{subfigure}
\caption{MSI auxiliary functions ($I < S < M$)}
\label{funcs}
\end{wrapfigure}
In Figure \ref{funcs} we define some auxiliary functions $maxCompatible(x)$ and $isModified(x)$ which are useful in describing protocols to maintain the invariant.

@@

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\uReq}{$c , a, x$}
    \If {$a$ is not present in $this$}
      \If {$this$ has no space for $a$}
        \State Choose address $a'$ for eviction;
        \State \textbf{forall} ($c' \in this.children$)
        \State \;\;\;\; \textbf{if} ($c'.state[a] > I$)
        \State \;\;\;\;\;\;\;\; \call{} $this.$\dReqL($c', a', I$);
        \State \textbf{if} ($this.state[a'] > I$)
        \State \;\;\;\; \call{} $this.$\dResp{}($this.parent, a', I$);
      \EndIf
      \State Allocate a line for $a$;
    \EndIf
    \State // $a$ is already present in $this$
    \State \textbf{forall} ($c' \in this.children$)
    \State \;\;\;\; \textbf{if} ($c'.state[a] > maxCompatible(x)$)
    \State \;\;\;\;\;\;\;\; \call{} $this.$\dReqL($c', a', maxCompatible(x)$);
    \State \call{} $this.$\uReqL{}($this.parent, a, x$);
  \State \textbf{if} ($c \neq core$)
  \State \;\;\;\; \call{} $this$.\uResp{}($c, a, x$);
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{$this$ cache's upgrade request handler -- request from one of $this$' children, $c$, to upgrade $c$'s state to $x$ for address $a$ }
\label{atomicUReq1}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\dReq}{$p , a, x$}
  \If {$this.state[a] > x$}
    \State \textbf{forall} ($c' \in this.children$)
    \State \;\;\;\; \textbf{if} (c'.state[a] > x)
    \State \;\;\;\;\;\;\;\; \call{} $this.$\dReqL($c', a, x$);
    \State \call{} $this.$\dResp($p, a, x$);
  \EndIf
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{$this$ cache's downgrade request handler -- request from $this$'
parent, $p$, to upgrade $p$'s state to $x$ for address $a$ }
\label{atomicDReq1}
\end{subfigure}

\caption{MSI protocol}
\label{msi-template}
\end{figure}

We will first describe the upgrade and downgrade methods as sequential procedures which can access and modify the state of any cache in the system (see
Figure \ref{msi-template}).  
When the upgrade request handler \uReq{} for a cache $this$ is invoked by a
cache $c$ to go to a particular state $x$, $this$ first makes sure that it has space to process it. This means that if $a$ is not present in $this$, it allocates a new line for that address, sometimes by evicting another line. This requires downgrading the status of the evicted address in all its children. Next, $this$ ensures that its
other children are ``compatible'' with $c$ being in state $x$ for the requested
address. For example, for $c$ to go to state $M$, all the children of $this$, except for $c$ must be in state $I$. If some children are not in state $I$ then the parent (\ie $this$) must downgrade them to go into state $I$. In the protocol, this is accomplished by the parent asking the child to downgrade their state to $maxCompatible(x)$. If the child's state is such that $isModified(state)$ is true, then it will contain the latest version of the data for address $a$, and hence this data has to be written back into the parent's cache.
If $this$' state is less than the requested state it invokes
its parent's handler for an upgrade request to $x$. Finally, when state of $this$ is $x$ and all its children are compatible with $x$, $this$ can upgrade
$c$'s state, and send the data if $c$'s state was $I$ previously. This is shown
in Figure \ref{atomicUReq1}.

Any cache except the LLC can get a downgrade request from its parent. Figure \ref{atomicDReq1} defines this downgrade method \dReq{} where the cache
ensures that each of its children's state is not higher than $x$, by invoking
their respective downgrade methods. When all the children have downgraded, the
cache downgrades its own state to $x$, and sends data back to the parent, if it
has been modified.

@@

\State // $this$ downgrades its own state and updates the parent's data
\Proc {\dResp}{$p, a, x$}
  \If {$isModified(c.state[a])$}
    \State $p.data[a] \gets this.data[a]$;
    \State // $p.data$ refers to $memory$ if $this$ is LLC
  \EndIf
  \State $this.state[a] \gets x$;
\EndProc

\State // In case of LLC, $this$ reads data from memory
\Proc {\sendMem{}}{$a$}
\EndProc
\end{algorithmic}

@@

The methods to handle upgrade and downgrade requests in turn use 5 auxiliary methods defined in Figure \ref{atomic}. These methods are very simple and could have been inlined. However, they do recursively call $this$' parent's \uReq{} or $this$' children's \uReq{}. We have chosen not to inline them because these functions encapsulate all accesses to other caches when we systematically transform this protocol for distributed implementation. Note however that \uReq{} and \dReq{} read the states of its children. We will make those reads local by keeping a directory as discussed in the next section.

>>>>

%\begin{wrapfigure}{r}{.6\linewidth}
\begin{figure}
\begin{subfigure}{.4\linewidth}
\centering
\begin{tabular}{|c|c|}
\hline
$M$ & $I$\\
$S$ & $S$\\
$I$ & $M$\\
\hline
\end{tabular}
\caption{$maxCompatible$}
\label{toCompat}
\end{subfigure}
\begin{subfigure}{.4\linewidth}
\centering
\begin{tabular}{|c|c|}
\hline
$M$ & $True$\\
$S$ & $False$\\
$I$ & $False$\\
\hline
\end{tabular}
\caption{$isModified$}
\label{isModified}
\end{subfigure}
\caption{MSI auxiliary functions ($I < S < M$)}
\label{funcs}
%\end{wrapfigure}
\end{figure}
In Figure \ref{funcs} we define some auxiliary functions $maxCompatible(x)$ and $isModified(x)$ which are useful in describing protocols to maintain the invariant.

@@

\begin{algorithmic}
\State //$this$ cache's upgrade request handler. Handles request from $c$ to upgrade $c$'s state to $x$ for address $a$
\Proc {\uReq}{$c , a, x$}
  \If {$x > c.state[a]$}
    \If {$a$ is not present in $this$}
      \If {$this$ has no space for $a$}
        \State Choose address $a'$ for eviction;
        \State \textbf{forall} ($c' \in this.children$)
        \State \;\;\;\; \textbf{if} ($c'.state[a] > I$)
        \State \;\;\;\;\;\;\;\; \call{} $this.$\dReqL($c', a', I$);
        \State \textbf{if} ($this.state[a'] > I$)
        \State \;\;\;\; \call{} $this.$\dResp{}($this.parent, a', I$);
      \EndIf
      \State Allocate a line for $a$;
    \EndIf
    \State // $a$ is already present in $this$
    \State \textbf{forall} ($c' \in this.children$)
    \State \;\;\;\; \textbf{if} ($c'.state[a] > maxCompatible(x)$)
    \State \;\;\;\;\;\;\;\; \call{} $this.$\dReqL($c', a', maxCompatible(x)$);
    \State \call{} $this.$\uReqL{}($this.parent, a, x$);
  \State \textbf{if} ($c \neq core$)
  \State \;\;\;\; \call{} $this$.\uResp{}($c, a, x$);
  \EndIf
\EndProc
\State //$this$ cache's downgrade request handler. Handles request from 
$p$ to downgrade $this$' state to $x$ for address $a$ 
\Proc {\dReq}{$p , a, x$}
  \If {$this.state[a] > x$}
    \State \textbf{forall} ($c' \in this.children$)
    \State \;\;\;\; \textbf{if} ($c'.state[a] > x$)
    \State \;\;\;\;\;\;\;\; \call{} $this.$\dReqL($c', a, x$);
    \State \call{} $this.$\dResp($p, a, x$);
  \EndIf
\EndProc
\end{algorithmic}
\caption{MSI protocol. In all the methods $p$ is the parent of $this$, $c$ is a child of $this$, and the methods work on address $a$ to change its state to $x$}
\label{msi-template}
\end{figure}

We will first describe the upgrade and downgrade methods as sequential procedures which can access and modify the state of any cache in the system (see
Figure \ref{msi-template}).  
When the upgrade request handler \uReq{} for a cache $this$ is invoked by a
cache $c$ to go to a particular state $x$, $this$ first makes sure that it has space to process it. This means that if $a$ is not present in $this$, it allocates a new line for that address, sometimes by evicting another line. This requires downgrading the status of the evicted address in all its children. Next, $this$ ensures that its
other children are ``compatible'' with $c$ being in state $x$ for the requested
address. For example, for $c$ to go to state $M$, all the children of $this$, except for $c$ must be in state $I$. If some children are not in state $I$ then the parent (\ie $this$) must downgrade them to go into state $I$. In the protocol, this is accomplished by the parent asking the child to downgrade their state to $maxCompatible(x)$. If the child's state is such that $isModified(state)$ is true, then it will contain the latest version of the data for address $a$, and hence this data has to be written back into the parent's cache.
If $this$' state is less than the requested state it invokes
its parent's handler for an upgrade request to $x$. Finally, when state of $this$ is $x$ and all its children are compatible with $x$, $this$ can upgrade
$c$'s state, and send the data if $c$'s state was $I$ previously. This is shown
in Figure \ref{msi-template}.

Any cache except the LLC can get a downgrade request from its parent. Figure \ref{msi-template} also defines this downgrade method \dReq{} where the cache
ensures that each of its children's state is not higher than $x$, by invoking
their respective downgrade methods. When all the children have downgraded, the
cache downgrades its own state to $x$, and sends data back to the parent, if it
has been modified.

@@

\State // $this$ downgrades its own state and updates the parent's data
\Proc {\dResp}{$p, a, x$}
  \If {$isModified(c.state[a])$}
    \State $p.data[a] \gets this.data[a]$;
    \State // $p.data$ refers to $memory$ if $this$ is LLC
  \EndIf
  \State $this.state[a] \gets x$;
\EndProc
\end{algorithmic}

@@

The methods to handle upgrade and downgrade requests in turn use four auxiliary methods defined in Figure \ref{atomic}. These methods are very simple and could have been inlined. However, they do recursively call $this$' parent's \uReq{} or $this$' children's \uReq{}. We have chosen not to inline them because these functions encapsulate all accesses to other caches when we systematically transform this protocol for distributed implementation. Note however that \uReq{} and \dReq{} read the states of its children. We will make those reads local by keeping a directory as discussed in the next section.
